---
title: "Severe Weather Events Analysis"
author: "José María Sebastián Carrillo"
date: "22 de febrero de 2018"
output:
  pdf_document:
    toc: true
    number_sections: true
  md_document:
    variant: markdown_github
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 "StormData.csv.bz2") [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "pd01016005curr.pdf")
* [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf "FNCDC_Storm_Events-FAQ_Page.pdf")

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


# Analysis

## Initial checks

First of all, we need to check the libraries present in the system:
```{r}
# Author <- "José Mª Sebastián Carrillo"

if (!require('plyr')) {
    stop('The package plyr was not installed!')
}

if (!require('dplyr')) {
    stop('The package dplyr was not installed!')
}

if (!require('ggplot2')) {
  stop('The package ggplot2 was not installed!')
}

if (!require('gridExtra')) {
  stop('The package gridExtra was not installed!')
}
```


And the, check if the raw data is in our system, and if the answer is not, download and uncompress it.

```{r}
currentFolder <- getwd()
dataFileCompressed <- "StormData.csv.bz2"

# Verify the file downloaded
if (!file.exists(dataFileCompressed)){
    dataFileCompressedUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(dataFileCompressedUrl, dataFileCompressed, method="curl")
}
```

## Data Processing

### Loading the data

Extract the data from the compressed file and store it in a variable, but only the columns that would apply to our analysis, in this case:

* STATE: State events took place
* EVTYPE: Event Type (e.g. tornado, flood, etc.)
* FATALITIES: Number of fatalities
* INJURIES: Number of injuries
* PROPDMG: Property damage estimates, entered as actual dollar amounts
* PROPDMGEXP: Alphabetic Codes to signify magnitude “K” for thousands, “M” for millions, and “B” for billions)
* CROPDMG: Crop damage estimates, entered as actual dollar amounts
* CROPDMGEXP: Alphabetic Codes to signify magnitude “K” for thousands, “M” for millions, and “B” for billions)
```{r cache = TRUE}
rawData <- read.csv(bzfile(dataFileCompressed))
workDataset <- rawData[,c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
str(workDataset)
```

### Transform the data

In this moment we can analyze the distinct types of event fatalities that we have in the dataset.
```{r}
# Show only a few for visualizatie more easily first time (document too long).
head(unique(workDataset$EVTYPE), n = 20L)
# If you want to show all the values, use the next instruction instead the above
# unique(workDataset$EVTYPE)
```

Now, we can aggregate the events types.
```{r cache = TRUE}
# Transforms (force) the data into text
workDataset$EVTYPE<-as.character(workDataset$EVTYPE)
# Change the literals
workDataset[grep("DROUGHT", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Drought"
workDataset[grep("URBAN|STREAM|FLOOD",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Flood"
workDataset[grep("HAIL", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Hail"
workDataset[grep("FOG", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Fog"
workDataset[grep("AVALANC", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Avalanche"
workDataset[grep("LIG", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Lightning"
workDataset[grep("FIRE", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Wild Fire"
workDataset[grep("WARM|HEAT", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Heat"
workDataset[grep("SLUMP|SLIDE", workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Landslide"
workDataset[grep("FUNNEL|DUST DEVIL|WHIRLWIND|LANDSPOUT|WATERSPOUT|TORN|NADO",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Tornado"
workDataset[grep("HURRICANE|TYPHOON|TROP",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Hurricane/Typhoon"
workDataset[grep("BEACH|SWELL|SEICHE|TSUNAMI|SEA|COAST|RIP|WAVE|SUR|TID",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Coastal Event"
workDataset[grep("HYPOTHERMIA|SLEET|GLAZE|LOW TEMPERATURE|FREEZ|FROST|SNOW|BLIZZARD|WINT|ICE|ICY|COLD",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Winter Event"
workDataset[grep("DAM|WET|MIX|PRECIPITATION|WATER|RAIN|SHOWER",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Rain"
workDataset[grep("DUST|WIND|BURST|HIGH|TSTM|THUN|APACHE COUNTY",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Thunderstorm/Wind"
workDataset[grep("OTHER|MARINE|SMOKE|VOLCANIC|TURBULENCE|DROWNING|\\?",
                 workDataset$EVTYPE,ignore.case = T),"EVTYPE"] <-"Other"
# Convert to factor
workDataset$EVTYPE<-factor(workDataset$EVTYPE)
# Show the changes
summary(workDataset$EVTYPE)
```

The next thing we should do is convert the numeric data codes (PROPDMGEXP and CROPDMGEXP), that multiplies the normal numeric values (PROPDMG and CROPDMG), to a normalized value in dollars.
For this, we need to know wich are the values
```{r}
table(workDataset$CROPDMGEXP)
table(workDataset$PROPDMGEXP)
```

Now we can construct a function that will make the "decode" this values to a numeric ones (in desc. order of appearances).
```{r}
decodeExponent <- function(value) {
    if (value == "") return(0)
    else if (value %in% c("k", "K"))
        return(3)
    else if (value %in% c("m", "M"))
        return(6)
    else if (!is.na(as.numeric(value))) 
        return(as.numeric(value))
    else if (value %in% c("b", "B"))
        return(9)
    else if (value %in% c("h", "H"))
        return(2)
    else if (value %in% c("-", "?", "+"))
        return(0)
    else return(0)
}
```

So we can use no the previous function to transform the data.
```{r cache = TRUE}
# Property damages
propExponent <- sapply(workDataset$PROPDMGEXP, FUN=decodeExponent)
workDataset$Property_Damage <- workDataset$PROPDMG * (10 ** propExponent)
# Crop damages
cropExponent <- sapply(workDataset$CROPDMGEXP, FUN=decodeExponent)
workDataset$Crop_Damage <- workDataset$CROPDMG * (10 ** cropExponent)
# Total damages
workDataset$Total_Damage <- workDataset$Property_Damage + workDataset$Crop_Damage
```

We start to construct the final dataset, contains the information that we'll show after, renaming the column names for more clear analysis.
```{r cache = TRUE}
weatherEvents <- workDataset[,c("STATE", "EVTYPE", "FATALITIES", "INJURIES",
                                "Property_Damage","Crop_Damage", "Total_Damage")]
names(weatherEvents)[names(weatherEvents) == 'STATE'] <- 'State'
names(weatherEvents)[names(weatherEvents) == 'EVTYPE'] <- 'Event_Type'
names(weatherEvents)[names(weatherEvents) == 'FATALITIES'] <- 'Fatalities'
names(weatherEvents)[names(weatherEvents) == 'INJURIES'] <- 'Injuries'
str(weatherEvents)
```

### Working with the data

Using this dataset, we can aggregate the data by the target columns, making the final datasets used for represntation.

First we summarize the health damages, sorting in descendant order, and removing the values that have not impact.
```{r cache = TRUE}
# Health fatalities
healthFatalities <- ddply(weatherEvents, .(Event_Type), summarize,
                          Fatalities = sum(Fatalities))
# Remove data with no impact
healthFatalities <- healthFatalities[(healthFatalities$Fatalities > 0), ]
# Order the final data
healthFatalitiesOrdered <- healthFatalities[order(healthFatalities$Fatalities,
                                                  decreasing = TRUE), ]
# Let's see the data
str(healthFatalitiesOrdered)
head(healthFatalitiesOrdered)

# Health injuries
healthInjuries <- ddply(weatherEvents, .(Event_Type), summarize,
                        Injuries = sum(Injuries))
# Remove data with no impact
healthInjuries <- healthInjuries[(healthInjuries$Injuries > 0), ]
# Order the final data
healthInjuriesOrdered <- healthInjuries[order(healthInjuries$Injuries,
                                              decreasing = TRUE), ]
# Let's see the data
str(healthInjuriesOrdered)
head(healthInjuriesOrdered)
```

And then, we summarize the property and crop damages, sorting in descendant order, and removing the values that have not impact.
```{r cache = TRUE}
# Property damages
propertyDamage <- ddply(weatherEvents, .(Event_Type), summarize,
                        Property_Damage = sum(Property_Damage))
# Remove data with no impact
propertyDamage <- propertyDamage[(propertyDamage$Property_Damage > 0), ]
# Order the final data
propertyDamageOrdered <- propertyDamage[order(propertyDamage$Property_Damage,
                                              decreasing = TRUE), ]
# Let's see the data
str(propertyDamageOrdered)
head(propertyDamageOrdered)

# Crop damages
cropDamage <- ddply(weatherEvents, .(Event_Type), summarize,
                    Crop_Damage = sum(Crop_Damage))
# Remove data with no impact
cropDamage <- cropDamage[(cropDamage$Crop_Damage > 0), ]
# Order the final data
cropDamageOrdered <- cropDamage[order(cropDamage$Crop_Damage,
                                              decreasing = TRUE), ]
# Let's see the data
str(cropDamageOrdered)
head(cropDamageOrdered)
```


## Results

Now we can make the first plot, that shows the influence in health of the top ten weather events.
```{r}
pHealthInjuries <- ggplot(data = head(healthInjuriesOrdered,10),
                          aes(x=reorder(Event_Type, Injuries), y=Injuries)) +
    geom_bar(stat="identity", fill="steelblue") +
    geom_text(aes(label = Injuries), hjust = -0.1, size = 3) +
    scale_y_continuous(limits=c(0, max(healthInjuriesOrdered$Injuries) * 1.05)) +
    coord_flip() +
    xlab("Event type") + ylab("Total number of injuries") +
    ggtitle("Top 10 weather events in the US - Health impact")

pHealthFatalities <- ggplot(data = head(healthFatalitiesOrdered,10),
                            aes(x=reorder(Event_Type, Fatalities), y=Fatalities)) +
    geom_bar(stat="identity", fill="olivedrab") +
    geom_text(aes(label = Fatalities), hjust = -0.1, size = 3) +
    scale_y_continuous(limits=c(0, max(healthFatalitiesOrdered$Fatalities) * 1.05)) +
    coord_flip() +
    xlab("Event type") + ylab("Total number of fatalities")

grid.arrange(pHealthInjuries, pHealthFatalities, nrow =2)

```

And the influence in the economics.
```{r}
pHealthInjuries <- ggplot(data = head(propertyDamageOrdered,10),
                          aes(x=reorder(Event_Type, Property_Damage),
                              y=round(Property_Damage/10**12,2))) +
    geom_bar(stat="identity", fill="steelblue") +
    geom_text(aes(label = round(Property_Damage/10**12,2)), hjust = -0.1, size = 3) +
    scale_y_continuous(limits=c(0, max(propertyDamageOrdered$Property_Damage/10**12) * 1.1)) +
    coord_flip() +
    xlab("Event type") + ylab("Property damage in trillion $ [10^12]") +
    ggtitle("Top 10 weather events in the US - Economic impact")

pHealthFatalities <- ggplot(data = head(cropDamageOrdered,10),
                            aes(x=reorder(Event_Type, Crop_Damage),
                                y=round(Crop_Damage/10**6,2))) +
    geom_bar(stat="identity", fill="olivedrab") +
    geom_text(aes(label = round(Crop_Damage/10**6,2)), hjust = -0.1, size = 3) +
    scale_y_continuous(limits=c(0, max(cropDamageOrdered$Crop_Damage/10**6) * 1.1)) +
    coord_flip() +
    xlab("Event type") + ylab("Crop damage in million $ [10^6]")

grid.arrange(pHealthInjuries, pHealthFatalities, nrow =2)

```
