---
title: "SevereWeatherEvents_Analysis"
author: "José María Sebastián Carrillo"
date: "22 de febrero de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 "StormData.csv.bz2") [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf "pd01016005curr.pdf")
* [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf "FNCDC_Storm_Events-FAQ_Page.pdf")

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


# Analysis

## Initial checks

First of all, we need to check the libraries present in the system:
```{r}
# Author <- "José Mª Sebastián Carrillo"

if (!require('plyr')) {
    stop('The package plyr was not installed!')
}
```


And the, check if the raw data is in our system, and if the answer is not, download and uncompress it.

```{r}
currentFolder <- getwd()
dataFileCompressed <- "StormData.csv.bz2"

# Verify the file downloaded
if (!file.exists(dataFileCompressed)){
    dataFileCompressedUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(dataFileCompressedUrl, dataFileCompressed, method="curl")
}
```

## Data Processing

### Loading the data

Extract the data from the compressed file and store it in a variable.
```{r cache = TRUE}
rawData <- read.csv(bzfile(dataFileCompressed))
str(rawData)
```
